Introduction to Lattice-based Attacks

### 2024-05-06 lattice_based_attacks.tex
version 2, fix grammar

\usepackage{alltt, fancyvrb, url}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{wrapfig}
% \usepackage{algorithm}% http://ctan.org/pkg/algorithms
% \usepackage[ruled,vlined]{algorithm2e}
% \usepackage{algorithm}
\usepackage[]{algorithm2e}
\usepackage{algorithmic}
\usepackage[utf8]{inputenc}
\usepackage{fontenc}
\usepackage{amsmath,stmaryrd,mathtools,algorithm}
\usepackage{amssymb}
%\usepackage{amsfonts}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{url}

@@

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{property}{Property}[section]
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\dd}{\cdot}

\title{Introduction to Lattice-based Attacks\\``Cryptography''}

\author{Di Santi Giovanni}
\date{\today}

@@

Lattices are powerful mathematical objects that can be used in computer science and mathematics to solves an extensive range of different problems.
In the recents years many post-quantum cryptographic protocols candidates involve the use of hard lattices problems,
for example the \textbf{Learning With Errors} \cite{LWE}.
Lattice constructions can also be used in cryptanalysis to break cryptographic schemes, typically using reduction
algorithms as \textbf{LLL} \cite{LLL}.

To introduce the mathematics of lattices a basic linear algebra background is needed, so the second chapter is a summary of the
basics properties needed to understand it. We used as main resource ``An Introduction to Mathematical Cryptography''\cite{mathcrypto14} to
explain them, but we also tried to gives some geometrical intuition behind some operations.

Cryptographic protocol as \textbf{RSA} and \textbf{ECDSA} are considered secure, however there are various attacks against a
relaxed-model of each of them, i.e., What if we know some bits of the secret key? What if we know a part of the message before the encryption?
We present an introduction to the attacks which involves lattices against these protocols. The code used to confirm the correctness
of the attacks is available on the repository \cite{repo} and the main reference was
``Recovering cryptographic keys from partial information, by example''\cite{cryptoeprint:2020:1506}.

@@

We'll start with some definitions and properties needed to understand the lattices, to move later on the mathematics behind them and some
hard problems related to it. At the end of the chapter the reader should be able to follow the math used in the other chapters.

@@

\begin{definition}
    \textbf{Vector space}.
\end{definition}
A \textit{vector space} $V$ is a subset of $\R^{m}$ which is closed under finite vector addition and scalar multiplication, with the property that

\begin{center}
   $a_1v_1 + a_2v_2 \in V$ for all $v_1,v_2 \in V$ and all $a_1,a_2 \in \R$
\end{center}

@@

Let $v_1,v_2,\ldots,v_k \in V$. A \textit{linear combination} of $v_1,v_2,\ldots,v_k \in V$ is any vector of the form

\begin{center}
    $\alpha_1v_1 + \alpha_2v_2 + \cdots + \alpha_kv_k$ with $\alpha_1, \ldots, \alpha_k \in \R$
\end{center}

\begin{definition}
    \textbf{Linear Independece}
\end{definition}

A set of vectors $v_1,v_2,\ldots,v_k \in V$ is \textit{linearly independent} if the the only way to get

@@

is to have $a_1 = a_2 = \cdots = a_k = 0$.

@@

Taken a set of linearly independent vectors $b = (v_1,\ldots,v_n) \in V$ we say that $b$ is a \textit{basis} of $V$ if $\forall w \in V$ we can write

@@

\begin{definition}
    \textbf{Vector's length}
\end{definition}

The vector's length or \textit{Euclidean norm} of $v = (x_1, x_2, \ldots, x_m)$ is

@@

@@

Let $v, w \in V \subset \R^m$ and $v = (x_1, x_2, \ldots, x_m), w = (y_1, y_2, \ldots, y_m)$, the \textit{dot product} of $v$ and $m$ is

\begin{center}
    $v \dd m = x_1y_1 + x_2y_2 + \cdots + x_my_m$

@@

    $v \dd m = \lVert v \rVert \lVert w \rVert \cos{\theta}$
\end{center}

where $\theta$ is the angle between $v$ and $w$ if we place the starting points of the vectors at the origin $O$.

@@

Geometrically speaking $v \cdot m$ is the length of $w$ projected to $v$ multiplied by the length of $v$ as shown in \ref{fig:dot_product}

\begin{definition}
    \textbf{Ortoghonal Basis} 
\end{definition}

An \textit{ortoghonal basis} for a vector space $V$ is a basis $v_1, \ldots, v_m$ with the property that

\begin{center}
    $v_i \dd v_j = 0$ for all $i \neq j$
\end{center}

If $\lVert v_i \rVert = 1$ for all $i$ then the basis is \textit{orthonormal}.

\begin{algorithm}[H]
    \vspace*{5px}
    $v_1^* = v1$\;
    \\
    \For{$i \gets 2$ \KwTo $n$}{
        $\mu_{ij} = v_i \cdot v_j^* / \lVert v_j^* \rVert ^ 2$ for $1 \le j < i$ \;
        \\
        $v_i^* = v_i - \sum_{j=1}^{i-1} \mu_{ij}v_j^*$ \;
    }
    \caption{Gram-Schmidt Algorithm}
    \label{alg:gram_schmidt}
\end{algorithm}

Let $b = (v_1, \ldots, v_n)$, be a basis for a vector space $V \subset \R^m$. There is an algorithm to create an orthogonal basis
$b^* = (v_1^*,\ldots,v_n^*)$.
The two bases have the property that Span$\{v_1,\ldots,v_i\}$ = Span$\{v_1^*,\ldots,v_i^*\}$ for all $i = 1,2,\ldots,n$

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.5\textwidth]{./img/gram_schmidt.png}
    \caption{Gram Schmidt orthogonalization}
    \label{fig:gram_schmidt}
\end{figure}

If we take $v_1=(4, 1), v_2=(2, 3)$ as basis and apply gram schmidt we obtain $u_1=v_1=(4, 1), u_2=(-10/17, 40/17)$ as shown in \ref{fig:gram_schmidt}

@@

The \textit{determinant} of a square matrix is a function that satisfies the following properties:

\begin{enumerate}
    \item The determinant is linear in each row.
    \item The determinant reverses sign if two rows are interchanged.
    \item The determinant of the identity matrix is equal to 1.
\end{enumerate}

If the determinant of a matrix is $0$ then the matrix is called \textbf{singular} (without an inverse).\\

Grant Sanderson created fantastic animations to understand the geometrical intuition behind the \textit{determinant}\cite{determinant}.

@@

Let $v_1,\ldots,v_n \in \R^m, m \ge n$ be linearly independent vectors. A \textit{Lattice} $L$ spanned by $\{v_1,\ldots,n_n\}$ is the set of 
all integer linear combinations of $v_1,\ldots,v_n$.

\begin{center}
    $L = \bigg\{\sum_{i=1}^{n} a_iv_i, a_i \in \Z \bigg\}$
\end{center}

If $v_i$ for every $i = 1,\ldots\,n$ has integer coordinates then the lattice is
called \textit{Integral Lattice}.

On the figure \ref{fig:lattice0} we show a lattice $L$ with bases $v=(3, 1)$ and $w=(-1, 1)$, and on \ref{fig:lattice1} the same lattice $L$ with
a different basis.

\begin{figure}[!tbp]
    \begin{minipage}[b]{0.50\textwidth}
        \includegraphics[width=\textwidth]{./img/lattice_b0.png}
        \caption{Lattice $L$ spanned by $v, w$}
        \label{fig:lattice0}
    \end{minipage}
    \hspace{\fill}
    \hspace{\fill}
    \hspace{\fill}
    \hspace{\fill}
    \hspace{\fill}
    \begin{minipage}[b]{0.50\textwidth}
        \includegraphics[width=\textwidth]{./img/lattice_b1.png}
        \caption{Lattice $L$ spanned by $v', w'$}
        \label{fig:lattice1}
    \end{minipage}
\end{figure}

@@

\textbf{The Shortest Vector Problem} (\texttt{SVP}): \textit{Find a nonzero vector $v \in L$ that minimez the Euclidean norm $\lVert v \rVert$.}

Gauss's developed an algorithm to find an optimal basis for a two-dimensional lattice given an arbitrary basis. The output of the algorithm
gives the shortest nonzero vector in $L$ and in this way solves the \texttt{SVP}.

@@

\begin{algorithm}[H]
    \vspace*{5px}
    \While{true}{
        \If{$\lVert v_2 \rVert < \lVert v_1 \rVert$}{
            swap $v_1, v_2$ \;
        }
        m = $\left \lfloor{(v_1 \cdot v_2) {\lVert v_1 \rVert}^{-2}} \right \rfloor$ \\
        \If{$m = 0$}{
            return $v_1, v_2$ \;
        }
        $v_2 = v_2 - mv_1$ \\
    }
    \caption{Gauss Basis Reduction}
\end{algorithm}

\textbf{Example}. Let $L$ a lattice spanned by $v_1 = (8, 4), v_2 = (7, 5)$, if we apply the gauss reduction algorithm we
obtain $w_1 = (1, -1), w_2 = (6, 6)$ \ref{fig:gauss_svp}. $w_1$ is the shortest nonzero vector in the lattice $L$.

The bigger the dimension of the lattice, the harder is the problem and there isn't a polynomial algorithm to solve it.

@@

\textbf{The Closest Vector Problem} (\texttt{CVP}): \textit{Given a vector $t \in \R^m$ that is not in $L$, find the vector $v \in L$
    closest to $t$, in other words find a vector $v \in L$ that minimizes the Euclidean norm $\lVert t - v \rVert$.}

@@

\texbtf{Example}. Let $L$ a lattice spanned by $v_1 = (0, 1), v_2 = (2, 0)$, and a target vector $t = (2.3, 2.4)$. The closest vector
to $t$ in the lattice $L$ is $w = (2, 2)$.\\

Both \textbf{CVP} and \textbf{SVP} are $\mathcal{NP}$-hard problems, thus the complexity of the problem increments with the dimension
of the lattice. They can be used as primitive to create a trapdoor functions for a cryptographic scheme.

There are various algorithms that can output a solution under certain constraints for an approximation of both \textbf{SVP}
and \textbf{CVP}, for example \textbf{LLL}\cite{LLL} and \textbf{Babai's algorithm}\cite{babai} respectively.

@@

The \textbf{Lenstra-Lenstra-Lovász} \textit{LLL}\cite{LLL} or \textit{$L^3$} is a polynomial time algorithm to find a ``short'' basis of a lattice.

@@

\textit{Let $L \in \Z^n$ be a lattice spanned by $B = \{v_1,\ldots,v_n\}$. The LLL algorithm outputs a reduced lattice
basis $\{w_1, \ldots, w_n\}$ with}

\begin{center}
    $\lVert w_i \rVert \le 2^{\frac{n(n-1)}{4(n-i+1)}} det(L)^{\frac{1}{n-i+1}}$ for $i=1,\ldots,n$
\end{center}

\textit{in time polynomial in n and in the bit-size of the entries of the basis matrix $B$}.\\

Basically the first vector of the new basis will be as short as possible, and the other will have increasing lengths. The new vectors
will be as orthogonal as possible to one another, i.e., the dot product $w_i \dd w_j$ will be close to zero.

@@

For example we can take the following basis (the rows are the vector) that span
a lattice $L$.

@@

Applying the LLL algorithm we obtain

@@

Where the first row is the shortest vector in the lattice $L$, and so solves the \textbf{SVP} problem.
For higher dimensions however the LLL algorithm outputs only an approximation for the \textbf{SVP} problem.

@@

\textbf{RSA} is one of the earliest and most used asymmetric cryptosystem.
The usual step to generate a public/private key for \textbf{RSA} is the following

>>>>

\usepackage{alltt, fancyvrb, url}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{wrapfig}
% \usepackage{algorithm}% http://ctan.org/pkg/algorithms
% \usepackage[ruled, vlined]{algorithm2e}
% \usepackage{algorithm}
\usepackage[]{algorithm2e}
\usepackage{titling}
\usepackage{algorithmic}
\usepackage[utf8]{inputenc}
\usepackage{fontenc}
\usepackage{amsmath, stmaryrd, mathtools, algorithm}
\usepackage{amssymb}
%\usepackage{amsfonts}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{url}

@@

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{property}{Property}[section]
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\dd}{\cdot}
\newcommand{\subtitle}[1]{
  \posttitle{
    \par\end{center}
    \begin{center}\large#1\end{center}
    \vskip0.5em}
}

\title{Introduction to Lattice-based Attacks}
\subtitle{Version 2}

\author{Giovanni Di Santi}
\date{}

@@

Lattices are potent mathematical structures widely utilized in computer science and mathematics to solve a broad spectrum of problems.
In recent years, numerous post-quantum cryptographic protocol candidates have employed the challenging problems associated with hard lattices,
for example, the \textbf{Learning With Errors} \cite{LWE}.
Additionally, lattice constructions are leveraged in cryptanalysis to compromise cryptographic schemes, typically through the use of
reduction algorithms such as \textbf{LLL} \cite{LLL}.

To comprehend the mathematics of lattices, a foundational understanding of linear algebra is essential. Therefore, the second chapter provides a summary of the
basic properties necessary for this understanding. We primarily referenced ``An Introduction to Mathematical Cryptography''\cite{mathcrypto14} to
elucidate these concepts, while also providing some geometric intuition behind certain operations.

Cryptographic protocols such as \textbf{RSA} and \textbf{ECDSA} are considered secure; however, various attacks can exploit a
relaxed model of each, e.g., what if some bits of the secret key are known? What if a portion of the message is known prior to encryption?
We introduce lattice-based attacks on these protocols and provide code in our repository \cite{repo} to verify the effectiveness of these attacks. The main reference for this discussion is
``Recovering cryptographic keys from partial information, by example''\cite{cryptoeprint:2020:1506}.

% Lattices are powerful mathematical objects that can be used in computer science and mathematics to solves an extensive range of different problems.
% In the recent years many post-quantum cryptographic protocols candidates involve the use of hard lattices problems,
% for example the \textbf{Learning With Errors} \cite{LWE}.
% Lattice constructions can also be used in cryptanalysis to break cryptographic schemes, typically using reduction
% algorithms as \textbf{LLL} \cite{LLL}.
% 
% To introduce the mathematics of lattices a basic linear algebra background is needed, so the second chapter is a summary of the
% basics properties needed to understand it. We used as main resource ``An Introduction to Mathematical Cryptography''\cite{mathcrypto14} to
% explain them, but we also tried to give some geometrical intuition behind some operations.
% 
% Cryptographic protocol as \textbf{RSA} and \textbf{ECDSA} are considered secure, however there are various attacks against a
% relaxed-model of each of them, i.e., What if we know some bits of the secret key? What if we know a part of the message before the encryption?
% We present an introduction to the attacks which involves lattices against these protocols. The code used to confirm the correctness
% of the attacks is available on the repository \cite{repo} and the main reference was
% ``Recovering cryptographic keys from partial information, by example''\cite{cryptoeprint:2020:1506}.

@@

This chapter introduces key definitions and properties essential for understanding lattices. It progresses to cover the mathematics of lattices and related complex problems. By the end of this chapter, the reader should be well-equipped to follow the mathematical discussions in subsequent chapters.

@@

\begin{definition}
    \textbf{Vector space}.
\end{definition}
A \textit{vector space} $V$ over $\R^{m}$ is a set closed under finite vector addition and scalar multiplication. This closure is characterized by:

\begin{center}
   $a_1v_1 + a_2v_2 \in V$ for any $v_1,v_2 \in V$ and $a_1,a_2 \in \R$
\end{center}

@@

Consider any vectors $v_1, v_2, \ldots, v_k \in V$. A \textit{linear combination} of these vectors is any vector expressed as:

\begin{center}
    $\alpha_1v_1 + \alpha_2v_2 + \cdots + \alpha_kv_k$ where $\alpha_1, \ldots, \alpha_k \in \R$
\end{center}

\begin{definition}
    \textbf{Linear Independence}
\end{definition}

A set of vectors $v_1, v_2, \ldots, v_k \in V$ is \textit{linearly independent} if the only solution to the equation

@@

is $a_1 = a_2 = \cdots = a_k = 0$.

@@

A set of linearly independent vectors $b = (v_1, \ldots, v_n) \in V$ constitutes a \textit{basis} of $V$ if every vector $w \in V$ can be represented as:

@@

\begin{definition}
    \textbf{Vector's Length}
\end{definition}

The \textit{Euclidean norm} or length of a vector $v = (x_1, x_2, \ldots, x_m)$ is given by:

@@

% \chapter{Linear Algebra Background}
% 
% We start with some definitions and properties needed to understand the lattices, to move later on the mathematics behind them and some
% hard problems related to it. At the end of the chapter the reader should be able to follow the math used in the other chapters.
% 
% \section{Vector Spaces}
% 
% \begin{definition}
%     \textbf{Vector space}.
% \end{definition}
% A \textit{vector space} $V$ is a subset of $\R^{m}$ which is closed under finite vector addition and scalar multiplication, with the property that
% 
% \begin{center}
%    $a_1v_1 + a_2v_2 \in V$ for all $v_1,v_2 \in V$ and all $a_1,a_2 \in \R$
% \end{center}
% 
% \begin{definition}
%     \textbf{Linear Combinations}
% \end{definition}
% 
% Let $v_1,v_2,\ldots,v_k \in V$. A \textit{linear combination} of $v_1,v_2,\ldots,v_k \in V$ is any vector of the form:
% 
% \begin{center}
%     $\alpha_1v_1 + \alpha_2v_2 + \cdots + \alpha_kv_k$ with $\alpha_1, \ldots, \alpha_k \in \R$
% \end{center}
% 
% \begin{definition}
%     \textbf{Linear Independece}
% \end{definition}
% 
% A set of vectors $v_1,v_2,\ldots,v_k \in V$ is \textit{linearly independent} if the the only way to get
% 
% \begin{center}
%     $a_1v_1 + a_2v_2 + \cdots + a_kv_k = 0$
% \end{center}
% 
% is to have $a_1 = a_2 = \cdots = a_k = 0$.
% 
% \begin{definition}
%     \textbf{Bases}
% \end{definition}
% 
% Taken a set of linearly independent vectors $b = (v_1,\ldots,v_n) \in V$ we say that $b$ is a \textit{basis} of $V$ if $\forall w \in V$ we can write
% 
% \begin{center}
%     $w = a_1v_1 + a_2v_2 + \cdots + a_nv_n$
% \end{center}
% 
% \begin{definition}
%     \textbf{Vector's length}
% \end{definition}
% 
% The vector's length or \textit{Euclidean norm} of $v = (x_1, x_2, \ldots, x_m)$ is
% 
% \begin{center}
%     $\lVert v \rVert = \sqrt{x_1^2 + x_2^2 + \cdots + x_m^2}$
% \end{center}

@@

Given vectors $v, w \in V \subset \R^m$ where $v = (x_1, x_2, \ldots, x_m)$ and $w = (y_1, y_2, \ldots, y_m)$, the \textit{dot product} of $v$ and $w$ is defined as:

\begin{center}
    $v \cdot w = x_1y_1 + x_2y_2 + \cdots + x_my_m$

@@

    $v \cdot w = \lVert v \rVert \lVert w \rVert \cos{\theta}$
\end{center}

where $\theta$ is the angle between $v$ and $w$ when their starting points are positioned at the origin $O$.

@@

Geometrically, $v \cdot w$ represents the length of the projection of $w$ onto $v$ multiplied by the length of $v$, as illustrated in \ref{fig:dot_product}.

\begin{definition}
    \textbf{Orthogonal Basis}
\end{definition}

An \textit{orthogonal basis} for a vector space $V$ consists of vectors $v_1, \ldots, v_m$ such that:

\begin{center}
    $v_i \cdot v_j = 0$ for all $i \neq j$
\end{center}

If $\lVert v_i \rVert = 1$ for each $i$, then the basis is termed \textit{orthonormal}.

\begin{algorithm}[H]
    \vspace*{5px}
    $v_1^* = v_1$\;
    \\
    \For{$i \gets 2$ \KwTo $n$}{
        $\mu_{ij} = v_i \cdot v_j^* / \lVert v_j^* \rVert ^ 2$ for $1 \le j < i$ \;
        \\
        $v_i^* = v_i - \sum_{j=1}^{i-1} \mu_{ij}v_j^*$ \;
    }
    \caption{Gram-Schmidt Algorithm}
    \label{alg:gram_schmidt}
\end{algorithm}

Given a basis $b = (v_1, \ldots, v_n)$ of a vector space $V \subset \R^m$, the Gram-Schmidt algorithm generates an orthogonal basis $b^* = (v_1^*,\ldots,v_n^*)$. These two bases satisfy the property that Span$\{v_1,\ldots,v_i\}$ = Span$\{v_1^*,\ldots,v_i^*\}$ for all $i = 1,2,\ldots,n$.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.5\textwidth]{./img/gram_schmidt.png}
    \caption{Gram Schmidt Orthogonalization}
    \label{fig:gram_schmidt}
\end{figure}

Using vectors $v_1=(4, 1)$ and $v_2=(2, 3)$ as a basis and applying the Gram-Schmidt process results in $u_1=v_1=(4, 1), u_2=(-10/17, 40/17)$, as depicted in \ref{fig:gram_schmidt}.

@@

The \textit{determinant} of a square matrix is a scalar value that characterizes the matrix and is defined by the following properties:

\begin{enumerate}
    \item The determinant is linear in each row.
    \item The determinant changes sign when two rows are swapped.
    \item The determinant of the identity matrix is 1.
\end{enumerate}

A matrix is termed \textbf{singular} if its determinant is $0$, indicating it lacks an inverse.\\

Animations created by Grant Sanderson provide a fantastic visual understanding of the geometric intuition behind the \textit{determinant}\cite{determinant}.

% \begin{definition}
%     \textbf{Dot Product}
% \end{definition}
% 
% Let $v, w \in V \subset \R^m$ and $v = (x_1, x_2, \ldots, x_m), w = (y_1, y_2, \ldots, y_m)$, the \textit{dot product} of $v$ and $m$ is
% 
% \begin{center}
%     $v \dd m = x_1y_1 + x_2y_2 + \cdots + x_my_m$
% 
%     or
% 
%     $v \dd m = \lVert v \rVert \lVert w \rVert \cos{\theta}$
% \end{center}
% 
% where $\theta$ is the angle between $v$ and $w$ if we place the starting points of the vectors at the origin $O$.
% 
% \begin{figure}[!b]
%     \centering
%     \includegraphics[scale=0.2]{./img/dot_product.png}
%     \caption{Dot Product By 3Blue1Brown \cite{3b1b}}
%     \label{fig:dot_product}
% \end{figure}
% 
% Geometrically speaking $v \cdot m$ is the length of $w$ projected to $v$ multiplied by the length of $v$ as shown in \ref{fig:dot_product}
% 
% \begin{definition}
%     \textbf{Ortoghonal Basis} 
% \end{definition}
% 
% An \textit{ortoghonal basis} for a vector space $V$ is a basis $v_1, \ldots, v_m$ with the property that
% 
% \begin{center}
%     $v_i \dd v_j = 0$ for all $i \neq j$
% \end{center}
% 
% If $\lVert v_i \rVert = 1$ for all $i$ then the basis is \textit{orthonormal}.
% 
% \begin{algorithm}[H]
%     \vspace*{5px}
%     $v_1^* = v1$\;
%     \\
%     \For{$i \gets 2$ \KwTo $n$}{
%         $\mu_{ij} = v_i \cdot v_j^* / \lVert v_j^* \rVert ^ 2$ for $1 \le j < i$ \;
%         \\
%         $v_i^* = v_i - \sum_{j=1}^{i-1} \mu_{ij}v_j^*$ \;
%     }
%     \caption{Gram-Schmidt Algorithm}
%     \label{alg:gram_schmidt}
% \end{algorithm}
% 
% Let $b = (v_1, \ldots, v_n)$, be a basis for a vector space $V \subset \R^m$. There is an algorithm to create an orthogonal basis
% $b^* = (v_1^*,\ldots,v_n^*)$.
% The two bases have the property that Span$\{v_1,\ldots,v_i\}$ = Span$\{v_1^*,\ldots,v_i^*\}$ for all $i = 1,2,\ldots,n$
% 
% \begin{figure}[htpb]
%     \centering
%     \includegraphics[width=0.5\textwidth]{./img/gram_schmidt.png}
%     \caption{Gram Schmidt orthogonalization}
%     \label{fig:gram_schmidt}
% \end{figure}
% 
% If we take $v_1=(4, 1), v_2=(2, 3)$ as basis and apply gram schmidt we obtain $u_1=v_1=(4, 1), u_2=(-10/17, 40/17)$ as shown in \ref{fig:gram_schmidt}
% 
% \begin{definition}
%     \textbf{Determinant}
% \end{definition}
% 
% The \textit{determinant} of a square matrix is a function that satisfies the following properties:
% 
% \begin{enumerate}
%     \item The determinant is linear in each row.
%     \item The determinant reverses sign if two rows are interchanged.
%     \item The determinant of the identity matrix is equal to 1.
% \end{enumerate}
% 
% If the determinant of a matrix is $0$ then the matrix is called \textbf{singular} (without an inverse).\\
% 
% Grant Sanderson created fantastic animations to understand the geometrical intuition behind the \textit{determinant}\cite{determinant}.

% \section{Lattices}
% 
% \begin{definition}
%     \textbf{Lattice}
% \end{definition}
% 
% Let $v_1,\ldots,v_n \in \R^m, m \ge n$ be linearly independent vectors. A \textit{Lattice} $L$ spanned by $\{v_1,\ldots,n_n\}$ is the set of 
% all integer linear combinations of $v_1,\ldots,v_n$.
% 
% \begin{center}
%     $L = \bigg\{\sum_{i=1}^{n} a_iv_i, a_i \in \Z \bigg\}$
% \end{center}
% 
% If $v_i$ for every $i = 1,\ldots\,n$ has integer coordinates then the lattice is
% called \textit{Integral Lattice}.
% 
% On the figure \ref{fig:lattice0} we show a lattice $L$ with bases $v=(3, 1)$ and $w=(-1, 1)$, and on \ref{fig:lattice1} the same lattice $L$ with
% a different basis.
% 
% \begin{figure}[!bp]
%     \begin{minipage}[b]{0.50\textwidth}
%         \includegraphics[width=\textwidth]{./img/lattice_b0.png}
%         \caption{Lattice $L$ spanned by $v, w$}
%         \label{fig:lattice0}
%     \end{minipage}
%     \hspace{\fill}
%     \hspace{\fill}
%     \hspace{\fill}
%     \hspace{\fill}
%     \hspace{\fill}
%     \begin{minipage}[b]{0.50\textwidth}
%         \includegraphics[width=\textwidth]{./img/lattice_b1.png}
%         \caption{Lattice $L$ spanned by $v', w'$}
%         \label{fig:lattice1}
%     \end{minipage}
% \end{figure}
% 
% \clearpage
% 
% \section{Lattice Problems}
% 
% \subsection{SVP}
% 
% \textbf{The Shortest Vector Problem} (\texttt{SVP}): \textit{Find a nonzero vector $v \in L$ that minimez the Euclidean norm $\lVert v \rVert$.}
% 
% Gauss's developed an algorithm to find an optimal basis for a two-dimensional lattice given an arbitrary basis. The output of the algorithm
% gives the shortest nonzero vector in $L$ and in this way solves the \texttt{SVP}.
% 
% \begin{figure}[!b]
%     \centering
%     \includegraphics[width=0.6\textwidth]{./img/gauss_svp.png}
%     \caption{Gauss reduction}
%     \label{fig:gauss_svp}
% \end{figure}
% 
% \begin{algorithm}[H]
%     \vspace*{5px}
%     \While{true}{
%         \If{$\lVert v_2 \rVert < \lVert v_1 \rVert$}{
%             swap $v_1, v_2$ \;
%         }
%         m = $\left \lfloor{(v_1 \cdot v_2) {\lVert v_1 \rVert}^{-2}} \right \rfloor$ \\
%         \If{$m = 0$}{
%             return $v_1, v_2$ \;
%         }
%         $v_2 = v_2 - mv_1$ \\
%     }
%     \caption{Gauss Basis Reduction}
% \end{algorithm}
% 
% \textbf{Example}. Let $L$ a lattice spanned by $v_1 = (8, 4), v_2 = (7, 5)$, if we apply the gauss reduction algorithm we
% obtain $w_1 = (1, -1), w_2 = (6, 6)$ \ref{fig:gauss_svp}. $w_1$ is the shortest nonzero vector in the lattice $L$.
% 
% The bigger the dimension of the lattice, the harder is the problem and there isn't a polynomial algorithm to solve it.
% 
% \subsection{CVP}
% 
% \textbf{The Closest Vector Problem} (\texttt{CVP}): \textit{Given a vector $t \in \R^m$ that is not in $L$, find the vector $v \in L$
%     closest to $t$, in other words find a vector $v \in L$ that minimizes the Euclidean norm $\lVert t - v \rVert$.}
% 
% \begin{figure}[!ht]
%     \centering
%     \includegraphics[width=0.6\textwidth]{./img/cvp.png}
%     \caption{CVP}
%     \label{fig:cvp}
% \end{figure}
% 
% \textbf{Example}. Let $L$ a lattice spanned by $v_1 = (0, 1), v_2 = (2, 0)$, and a target vector $t = (2.3, 2.4)$. The closest vector
% to $t$ in the lattice $L$ is $w = (2, 2)$.\\
% 
% Both \textbf{CVP} and \textbf{SVP} are $\mathcal{NP}$-hard problems, thus the complexity of the problem increments with the dimension
% of the lattice. They can be used as primitive to create a trapdoor functions for a cryptographic scheme.
% 
% There are various algorithms that can output a solution under certain constraints for an approximation of both \textbf{SVP}
% and \textbf{CVP}, for example \textbf{LLL}\cite{LLL} and \textbf{Babai's algorithm}\cite{babai} respectively.

@@

Consider $v_1, \ldots, v_n \in \R^m$ with $m \ge n$, which are linearly independent vectors. A \textit{lattice} $L$, spanned by $\{v_1, \ldots, v_n\}$, comprises all integer linear combinations of these vectors:

\begin{center}
    $L = \bigg\{\sum_{i=1}^{n} a_iv_i , a_i \in \Z \bigg\}$
\end{center}

If every $v_i$ ($i = 1, \ldots, n$) has integer coordinates, the lattice is termed an \textit{integral lattice}.

Figures \ref{fig:lattice0} and \ref{fig:lattice1} depict the lattice $L$ with bases $v=(3, 1)$ and $w=(-1, 1)$, and another basis configuration, respectively.

\begin{figure}[H]
    \begin{minipage}[b]{0.50\textwidth}
        \includegraphics[width=\textwidth]{./img/lattice_b0.png}
        \caption{Lattice $L$ spanned by $v, w$}
        \label{fig:lattice0}
    \end{minipage}
    \hspace{\fill}
    \begin{minipage}[b]{0.50\textwidth}
        \includegraphics[width=\textwidth]{./img/lattice_b1.png}
        \caption{Lattice $L$ spanned by $v', w'$}
        \label{fig:lattice1}
    \end{minipage}
\end{figure}

@@

\textbf{The Shortest Vector Problem} (\texttt{SVP}): \textit{Find a nonzero vector $v \in L$ that minimizes the Euclidean norm $\lVert v \rVert$.}

Gauss developed an algorithm to determine an optimal basis for a two-dimensional lattice from any given basis. This method outputs the shortest nonzero vector in $L$, thereby solving the \texttt{SVP}.

@@

\begin{algorithm}[H]
    \vspace*{5px}
    \While{true}{
        \If{$\lVert v_2 \rVert < \lVert v_1 \rVert$}{
            swap $v_1, v_2$\;
        }
        m = $\left \lfloor{(v_1 \cdot v_2) {\lVert v_1 \rVert}^{-2}} \right \rfloor$\;
        \If{$m = 0$}{
            return $v_1, v_2$\;
        }
        $v_2 = v_2 - mv_1$\;
    }
    \caption{Gauss Basis Reduction}
\end{algorithm}

\textbf{Example}: In a lattice $L$ spanned by $v_1 = (8, 4), v_2 = (7, 5)$, applying the Gauss reduction algorithm results in $w_1 = (1, -1), w_2 = (6, 6)$ as shown in \ref{fig:gauss_svp}. $w_1$ represents the shortest nonzero vector in lattice $L$.

The complexity of solving \texttt{SVP} increases with the dimension of the lattice, making it a challenging problem without a known polynomial-time solution.

@@

\textbf{The Closest Vector Problem} (\texttt{CVP}): \textit{Given a vector $t \in \R^m$ that is not in $L$, find the vector $v \in L$ closest to $t$, minimizing the Euclidean norm $\lVert t - v \rVert$.}

@@

\textbf{Example}: For a lattice $L$ spanned by $v_1 = (0, 1), v_2 = (2, 0)$, with target vector $t = (2.3, 2.4)$, the closest vector in the lattice $L$ is $w = (2, 2)$, as depicted in \ref{fig:cvp}.

Both \textbf{CVP} and \textbf{SVP} are considered $\mathcal{NP}$-hard problems, and their complexity increases with the dimension of the lattice. They serve as primitives for creating trapdoor functions in cryptographic schemes.

Several algorithms can approximate solutions for both \textbf{SVP} and \textbf{CVP}, such as \textbf{LLL}\cite{LLL} and \textbf{Babai's algorithm}\cite{babai} respectively.

@@

The \textbf{Lenstra-Lenstra-Lovász (LLL)} algorithm is a polynomial-time method designed to compute a "short" basis for a lattice.

@@

Given a lattice $L \in \Z^n$ spanned by $B = \{v_1,\ldots,v_n\}$, the LLL algorithm produces a reduced basis $\{w_1, \ldots, w_n\}$ such that:

\begin{center}
    $\lVert w_i \rVert \le 2^{\frac{n(n-1)}{4(n-i+1)}} \cdot det(L)^{\frac{1}{n-i+1}}$ for $i=1,\ldots,n$
\end{center}

This process is executed in time polynomial in $n$ and the bit-size of the entries in the basis matrix $B$.\\

In this reduced basis, the vectors are as short as possible starting from the first, with each subsequent vector being longer. Additionally, these vectors are almost orthogonal to each other, meaning the dot product $w_i \cdot w_j$ is close to zero.

@@

Consider the following basis, represented in matrix form, that spans a lattice $L$:

@@

After applying the LLL algorithm, the basis is transformed to:

@@

Here, the first row is the shortest vector in the lattice $L$, thus addressing the \textbf{SVP} problem. For higher dimensions, however, the LLL algorithm provides only an approximation to the \textbf{SVP}.

@@

\textbf{RSA} is one of the earliest and most used asymmetric cryptosystem.
The steps to generate a public/private key for \textbf{RSA} are:

\begin{enumerate}
    \item Fix $e = 65537$ or $e = 3$ (public).
    \item Find two primes $p, q$ such that $p - 1$ and $q - 1$ are relatively prime to $e$, gcd$(e, p-1) = 1$ and gcd$(e, q-1) = 1$.
    \item Compute $N = p * q$ and $\phi(n) = (p-1) * (q-1)$
    \item Calculate $d$ (private) as the multiplicative inverse of $e$ modulo $\phi(n)$.
    \item $(N, e)$ is the public key, $(N, d)$ is the private key.
\end{enumerate}

To encrypt a message $m$ with \textbf{textbook RSA}:
